---
title: "linearReqgression"
author: "Sherif Shawashen"
date: "2024-10-20"
output: html_document
---

# functions

```{r }
 library(ggplot2)
library(ggfortify)


#visualize the residuals
visualize_residuals <- function(dependent,dependent_name, independent,independent_name, data_model) {
  
  # Get predicted values
  WHO$predicted <- predict(data_model)
  
  #
  ggplot(WHO, aes(x = independent, y = dependent)) +
    geom_point() +
    geom_segment(aes (xend = independent, yend = predicted), color = "red") + 
    geom_smooth(method = "lm", se = FALSE, color = "blue") +  
    theme_minimal() + 
    labs(title = paste("Visualizing Residuals: ", dependent_name, " vs ", independent_name),
         x = independent_name,
         y = dependent_name)
}


#visualize the linear reggression 
visualize_linear_reggression <- function(dependent,dependent_name, independent,independent_name) {
  
  ggplot(WHO, aes(x = independent, y = dependent)) +
    geom_point() +
    geom_smooth(method = "lm", se = FALSE, color = "blue") +  
    theme_minimal() + 
    labs(title = paste("Visualizing linear reggression: ", dependent_name, " vs ", independent_name),
         x = independent_name,
         y = dependent_name)
}


visualize_confidence_interval <- function(function_model,dependent,dependent_name, independent,independent_name)
{
 

# create new data for predictions
new_data <- data.frame(alcohol = seq(min(independent), max(independent), length.out = 200))

CI_95 <- predict(function_model, newdata = new_data, interval = "confidence", level = 0.95)

new_data$fit <- CI_95[, "fit"]
new_data$up <- CI_95[, "upr"]
new_data$down <- CI_95[, "lwr"]


ggplot() +  
  geom_smooth(method = "lm", se = FALSE, color = "blue") + 
  geom_line(data = new_data, aes(x = alcohol, y = fit), color = "red") +  # Regression line
  
  geom_line(data = new_data, aes(x = alcohol, y = up), linetype = "dashed", size = 1, color = "green") + 
  
  geom_line(data = new_data, aes(x = alcohol, y = down), linetype = "dashed", size = 1, color = "green") +  
  
  theme_minimal() +
  labs(title = paste("Linear Regression with 95% Confidence Interval for ",independent_name , " vs ",  dependent_name),
       x = independent_name,
       y = dependent_name)  


}

```


# function to change the names of data fram to lowercase and "." to "_" so we can access it easily
# link for gsub=> https://stackoverflow.com/questions/11776287/remove-pattern-from-string-with-gsub
``` {r}


WHO <- read.csv("Life Expectancy Data.csv")

LowerAndDotReplace <- function(data_frame_pass){
      # convert to lowercase
      names(data_frame_pass) <- tolower(names(data_frame_pass))
      # conver . with _ 
      names(data_frame_pass) <- gsub("\\.", "_", names(data_frame_pass))

  return(data_frame_pass)
}

WHO <-na.omit(WHO)
# class the function

WHO <- LowerAndDotReplace(WHO)
names(WHO)

```



## What is the relationship between consumption of alcohol and life expectancy?

```{r }

library(ggplot2)
library(dplyr)
library(MASS)

# Linear Model
life_expectancy_alcohol_model <- lm(life_expectancy ~ alcohol, data=WHO)

summary(life_expectancy_alcohol_model)

```

Expalin the output:- 

# Coefficients:
1- (Intercept): The model predicts that when the alcohol consumption is zero the, life expectancy is 64.76 years.

2- alcohol: the model says that for each unit increase in alcohol consumption, life expectancy increases by 0.87925  years, holding all other factors are constant.

3- p values : the models says that the p values for both Intercept and alcohol are less than 5% which indicats a realtionship between the alcohol consumption and life expectancy and we can reject null hypothesis.

4- Both the intercept and slope have very small standard errors, indicating the coefficient estimates are precise.

# Residual standard error:
 1- the model says that Residual standard error is 8.054 which is a big number that indicates the spread of the Residual in the model.
 2- the model says that the RES on 1647 degrees of freedom which means that the model has 1647 data point left after fitting the model
 
 
# R-squared
the model Multiple R-squared is 0.1622 which means that the model only explians 16% life expectancy so the model is not a good one.


# trying to overfit the model using polynomial

```{r }

overfitmo_del <- lm(life_expectancy ~ poly(alcohol, 2), data=WHO)
summary(overfitmo_del)


```
after trying to overfit the model the  R-squared increases to .17 which is not effecient as well.




#visualzing alcohol consumption vs life expectancy model    
```{r }
 
 visualize_linear_reggression(WHO$life_expectancy,"life_expectancy", WHO$alcohol,"alcohol")

```


# Visualizing residuals
```{r }
  #Predict life expectancy based on alcohol consumption using the linear model and
  #add it as a column in the data fram 
  WHO$predicted_life <- predict(life_expectancy_alcohol_model)
  visualize_residuals(WHO$life_expectancy,"life_expectancy",WHO$alcohol,"alcohol",life_expectancy_alcohol_model)
```



What I can notice form the graph is the the resduals are not normally distributed




# confidence interval for the model
```{r}
confint(life_expectancy_alcohol_model)

```

1- we can see that confidence interval for **intercept** is between 64.7308102 and 65.9022146 which mena that 95% if of the Intercept lies in this range.
2- the same for  confidence interval for **alcohol** whic is between 0.7826617 and  0.9758293 , meaning that 95% if of the alcohol lies in this range. 
3- the confidence interval renge for booth **intercept**  and **alcohol** does not contain a zero which impies a that we can reject the null hypothesis and there is a statistically significant relationship between alcohol consumption and life expectancy.

#visualzing confidence interval

```{r}

visualize_confidence_interval(life_expectancy_alcohol_model, 
                              WHO$life_expectancy, 
                              "life expectancy", 
                              WHO$alcohol, 
                              "alcohol consumption")


```


