---
title: "linearReqgression"
author: "Sherif Shawashen"
date: "2024-10-20"
output: html_document
---

# functions

```{r }
 library(ggplot2)
library(ggfortify)


#visualize the residuals
visualize_residuals <- function(dependent,dependent_name, independent,independent_name, data_model) {
  
  # Get predicted values
  WHO$predicted <- predict(data_model)
  
  #
  ggplot(WHO, aes(x = independent, y = dependent)) +
    geom_point() +
    geom_segment(aes (xend = independent, yend = predicted), color = "red") + 
    geom_smooth(method = "lm", se = FALSE, color = "blue") +  
    theme_minimal() + 
    labs(title = paste("Visualizing Residuals: ", dependent_name, " vs ", independent_name),
         x = independent_name,
         y = dependent_name)
}


#visualize the linear reggression 
visualize_linear_reggression <- function(dependent,dependent_name, independent,independent_name) {
  
  ggplot(WHO, aes(x = independent, y = dependent)) +
    geom_point() +
    geom_smooth(method = "lm", se = FALSE, color = "blue") +  
    theme_minimal() + 
    labs(title = paste("Visualizing linear reggression: ", dependent_name, " vs ", independent_name),
         x = independent_name,
         y = dependent_name)
}


visualize_confidence_interval <- function(function_model,dependent,dependent_name, independent,independent_name)
{
 

# create new data for predictions
new_data <- data.frame(alcohol = seq(min(independent), max(independent), length.out = 200))

CI_95 <- predict(function_model, newdata = new_data, interval = "confidence", level = 0.95)

new_data$fit <- CI_95[, "fit"]
new_data$up <- CI_95[, "upr"]
new_data$down <- CI_95[, "lwr"]


ggplot() +  
  geom_smooth(method = "lm", se = FALSE, color = "blue") + 
  geom_line(data = new_data, aes(x = alcohol, y = fit), color = "red") +  # Regression line
  
  geom_line(data = new_data, aes(x = alcohol, y = up), linetype = "dashed", size = 1, color = "green") + 
  
  geom_line(data = new_data, aes(x = alcohol, y = down), linetype = "dashed", size = 1, color = "green") +  
  
  theme_minimal() +
  labs(title = paste("Linear Regression with 95% Confidence Interval for ",independent_name , " vs ",  dependent_name),
       x = independent_name,
       y = dependent_name)  


}


# coefficient and their confidence interval 
library(ggplot2)

visualize_coefficient_CI <- function(function_model) {
  # Create data frame for coefficients
  df_coefficient <- as.data.frame(confint(function_model))
  df_coefficient$coefficient <- rownames(df_coefficient)  # Corrected spelling
  df_coefficient$estimate <- coef(function_model)
  
  # Plotting
  ggplot(df_coefficient, aes(x = coefficient, y = estimate)) +  # Corrected aes mapping
    geom_point(size = 3) +
    geom_errorbar(aes(ymin = `2.5 %`, ymax = `97.5 %`), width = 0.2) +
    labs(title = "Coefficients and Their Confidence Intervals", 
         x = "Coefficient", 
         y = "Estimate") +  # Corrected spelling
    theme_minimal() 
}


```


# function to change the names of data fram to lowercase and "." to "_" so we can access it easily
# link for gsub=> https://stackoverflow.com/questions/11776287/remove-pattern-from-string-with-gsub
``` {r}


WHO <- read.csv("Life Expectancy Data.csv")

LowerAndDotReplace <- function(data_frame_pass){
      # convert to lowercase
      names(data_frame_pass) <- tolower(names(data_frame_pass))
      # conver . with _ 
      names(data_frame_pass) <- gsub("\\.", "_", names(data_frame_pass))

  return(data_frame_pass)
}

WHO <-na.omit(WHO)
# class the function

WHO <- LowerAndDotReplace(WHO)
names(WHO)

```



##  Question 1 : What is the relationship between consumption of alcohol and life expectancy?

```{r }

library(ggplot2)
library(dplyr)
library(MASS)

# Linear Model
life_expectancy_alcohol_model <- lm(life_expectancy ~ alcohol, data=WHO)

summary(life_expectancy_alcohol_model)

```

Expalin the output:- 

# Coefficients:
1- (Intercept): The model predicts that when the alcohol consumption is zero the, life expectancy is 64.76 years.

2- alcohol: the model says that for each unit increase in alcohol consumption, life expectancy increases by 0.87925  years, holding all other factors are constant.

3- p values : the models says that the p values for both Intercept and alcohol are less than 5% which indicats a realtionship between the alcohol consumption and life expectancy and we can reject null hypothesis.

4- Both the intercept and slope have very small standard errors, indicating the coefficient estimates are precise.

# Residual standard error:
 1- the model says that Residual standard error is 8.054 which is a big number that indicates the spread of the Residual in the model.
 2- the model says that the RES on 1647 degrees of freedom which means that the model has 1647 data point left after fitting the model
 
 
# R-squared
the model Multiple R-squared is 0.1622 which means that the model only explians 16% life expectancy so the model is not a good one.


# trying to overfit the model using polynomial

```{r }

overfitmo_del <- lm(life_expectancy ~ poly(alcohol, 2), data=WHO)
summary(overfitmo_del)


```
after trying to overfit the model the  R-squared increases to .17 which is not effecient as well.




#visualzing alcohol consumption vs life expectancy model    
```{r }
 
 visualize_linear_reggression(WHO$life_expectancy,"life_expectancy", WHO$alcohol,"alcohol")

```


# Visualizing residuals
```{r }
  #Predict life expectancy based on alcohol consumption using the linear model and
  #add it as a column in the data fram 
  WHO$predicted_life <- predict(life_expectancy_alcohol_model)
  visualize_residuals(WHO$life_expectancy,"life_expectancy",WHO$alcohol,"alcohol",life_expectancy_alcohol_model)
```



What I can notice from the graph is the the resduals are not normally distributed




# confidence interval for the model
```{r}
confint(life_expectancy_alcohol_model)


#visualzing confidence interval
visualize_coefficient_CI(life_expectancy_alcohol_model)

```




1- we can see that confidence interval for **intercept** is between 64.7308102 and 65.9022146 which mena that 95% if of the Intercept lies in this range.
2- the same for  confidence interval for **alcohol** whic is between 0.7826617 and  0.9758293 , meaning that 95% if of the alcohol lies in this range. 
3- the confidence interval renge for booth **intercept**  and **alcohol** does not contain a zero which impies a that we can reject the null hypothesis and there is a statistically significant relationship between alcohol consumption and life expectancy.

#visualzing confidence interval

```{r}

visualize_confidence_interval(life_expectancy_alcohol_model, 
                              WHO$life_expectancy, 
                              "life expectancy", 
                              WHO$alcohol, 
                              "alcohol consumption")


```



# Question 2 : What is the relationship between the number of years in schooling and GDP per capita?

```{r }


library(ggplot2)
library(dplyr)
library(MASS)

# Linear Model
gdp_schooling_model <- lm(gdp ~ schooling, data=WHO)

summary(gdp_schooling_model)

```

#coefficients:
1- (Intercept): The model predicts that when schooling is zero the GDP is -17717.0 unit.

2- schooling: The output shows  that for each unit increase in schooling, GDP increases by 1921.1 units, holding all other factors constant.

3- p values: The output shows  that the p values for both the Intercept and schooling are less than 5%, which indicates a relationship between schooling and GDP, and we can reject the null hypothesis.

4- Both the intercept and slope have big standard errors, especially **Intercept** which is 1111.9 , indicating the coefficient estimates are not precise.

#Residual standard error:
1- The output shows that the Residual standard error is 10140  which is a significant number that indicates the spread of the residuals in the model.

2- The output shows  that there are 1647 degrees of freedom, which means that the model has 1647 data points left after fitting the model.

#R-squared:
The model's Multiple R-squared is 0.219, which means that the model only explains 21.9% of the variance in GDP which means that the model is not strong.

#visualzing GDP vs schooling model    
```{r }
 
 visualize_linear_reggression(WHO$gdp,"GDP", WHO$schooling,"schooling")

```

# Visualizing residuals
```{r }
  #Predict GDP based on Schooling using the linear model and
  #add it as a column in the data fram 
  WHO$predicted_GDP <- predict(gdp_schooling_model)
  visualize_residuals(WHO$gdp,"GDP",WHO$schooling,"Schooling",gdp_schooling_model)
```

# confidence interval for coefficient for the model
```{r}
confint(gdp_schooling_model)


#visualzing confidence interval
visualize_coefficient_CI(gdp_schooling_model)

```

1- We can see that the confidence interval for the **intercept** ranges from -19897.975 to -15536.084. This means that we can be 95% confident that the true value of the intercept lies within this range.

Similarly, the confidence interval for schooling is between 1745.715 and 2096.409, indicating that we can be 95% confident that the true effect of schooling on GDP falls within this interval.

Since the confidence intervals for both the intercept and schooling do not include zero, we can reject the null hypothesis. which suggests that there is a statistically significant relationship between schooling and GDP.



#visualzing confidence interval for the linear Reqression

```{r}

visualize_confidence_interval(life_expectancy_alcohol_model, 
                              WHO$GDP, 
                              "GDP", 
                              WHO$schooling, 
                              "schooling")

```
