---
title: "Mulit"
author: "Sherif Shawashen"
date: "2024-10-21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## load the dataset and clean the data using omit 
# function to change the names of data fram to lowercase and "." to "_" so we can access it easily
# link for gsub=> https://stackoverflow.com/questions/11776287/remove-pattern-from-string-with-gsub

```{r cars}
WHO <- read.csv("Life Expectancy Data.csv")


# function to function to change the names of data fram to lowercase and "." to "_" so we can access it easily
LowerAndDotReplace <- function(data_frame_pass){
      # convert to lowercase
      names(data_frame_pass) <- tolower(names(data_frame_pass))
      # conver . with _ 
      names(data_frame_pass) <- gsub("\\.", "_", names(data_frame_pass))

  return(data_frame_pass)
}
# removing na values
WHO <-na.omit(WHO)

WHO <- LowerAndDotReplace(WHO)

class
names(WHO)
```

##  Does the country’s GDP, Schooling or Total expenditure affect life expectancy?
```{r}
#check the class of the status column
class(WHO$status)
# check if there is other values rather than "Developing" "Developed" 
unique(WHO$status)

```


```{r}

# create a multilinear regression 
multi_model_gdp_status <- lm(life_expectancy ~ gdp + status + schooling + total_expenditure , data = WHO)
summary(multi_model_gdp_status)

```
# coefficients:

1- (**Intercept**): The model predicts that when all variables GDP, schooling, and total expenditure are zero and statusis Developed (baseline) , life expectancy equals 45.67 years. and for **intercept** The p-value is less than 5% , which indicates that this estimate is statistically significant.

2- GDP: The model shows that for each unit increase in GDP, life expectancy increases by 8.576e-05 years, holding all other factors constant and status is Developed (baseline). The  p-value less than 5% indicates a statistically significant relationship between GDP and life expectancy and we can reject null hypothesis.

3- status (Developing) -1.465e+00 :  is This is the difference in the intercept between "Developed" and "Developing", holding all other variables GDP, schooling, and total expenditure zero. The p-value is 0.0041, which shows this estimate is statistically significant.

schooling: The model shows that for each unit increase in schooling, life expectancy increases by 2.041e+00 years, holding all other factors constant and status = "Developed" . The v p-value (less than %5) indicates a statistically significant relationship between life expectency and Schooling.

total_expenditure: The model Shows that for each unit increase in total expenditure, the life expectancy decreases by 5.722e-02 years. However, the p-value of 0.3857 higher than 5% which means  that this effect is not statistically significant we can not reject null hypothesis.

#Residual standard error:
1- Residual standard error is 5.942 This is a measure of the variability of the residuals, which means that the Residual are not spread out in the model
2- Multiple R-squared: 0.5449 This means that the model explains 54.49% of the variance in life expectency, wich is not very high


#R- Squared

the model shows that Multiple R-squared = 0.5449 which means the variance in the life_expectancy that is explained by the independent variables (gdp, total_expenditure....) is  54.49% of the variability in life expectancy can be explained by the model. and Adjusted R-squaredis different from Multiple R-squared because the model have more than one predictor and Adjusted R-squared of 54.38% indicates that after accounting for the number of predictors, the model still explains about 54.38% of the variance in life expectancy which is less than Multiple R-squared which means the model includes unnecessary predictors.


# Multicollinearity for the multi_model_gdp_status model

```{r}

# show the corelation in the model
cor( WHO[,c("gdp" ,"schooling" , "total_expenditure")])


```
1- From the correlation matrix above, we can see that GDP and schooling are correlated, with a correlation estimate of 0.4679470. Therefore, we will remove one of them. Based on the correlation estimate, the relationship between total expenditure and GDP is less than that of schooling and total expenditure, so we will remove schooling from the model.


#multi_model_gdp_status after removing corelated factors

```{r}
# creating the model after removing correlated predictors
multi_model_gdp_status <- lm(life_expectancy ~ gdp + total_expenditure , data = WHO)

summary(multi_model_gdp_status)
```
#coeffectient

**Intercept**: The model predicts that when all variables (GDP and total expenditure) are zero, life expectancy equals 6.525e+01  years. The p-value is less than 5%, which indicates that this estimate is statistically significant.

**GDP**: The model shows that for each unit increase in GDP, life expectancy increases by 3.247e-04 years, holding all other factors constant. The p-value is less than 5%, indicating a statistically significant relationship between GDP and life expectancy, allowing us to reject the null hypothesis.

**Total Expenditure**: The model indicates that for each unit increase in total expenditure, life expectancy increases by 3.761e-01 years, holding all other factors constant. The p-value of 1.16e-05 is less than 5%, which means that this estimate is statistically significant, allowing us to reject the null hypothesis.

#R-Squared

1- The model shows that the Multiple R-squared is 0.2041 which means that the variance in life expectancy explained by the independent variables (GDP and total expenditure) is 20.41%. which is not a big number. The Adjusted R-squared of 0.2031 is slightly lower, indicating that after accounting for the number of predictors, the model explains about 20.31% of the variance in life expectancy. 

# Interactions Between  GDP and total_expenditure

```{r}

# Call interaction_model_GDP_total_expenditure function to see the interaction between GDP and total_expenditure
interaction_model_GDP_total_expenditure <- lm(WHO$life_expectancy ~ WHO$total_expenditure * WHO$gdp)
summary(interaction_model_GDP_total_expenditure)

```
# coefficients:

1- (**Intercept**): The model predicts that when all variables GDP, and total expenditure are zero  , life expectancy equals 6.435e+01 years. and for **intercept** The p-value is less than 5% , which indicates that this estimate is statistically significant.

2- **GDP**: The model shows that for each unit increase in GDP, life expectancy increases by5.230e-01 years, holding all other factors constant. The  p-value less than 5% indicates a statistically significant relationship between GDP and life expectancy and we can reject null hypothesis.


**WHO$total_expenditure:WHO$gdp** : The model shows that the effect of gdp on the life expectency changes by -1.795e-05 when total_expenditure increases by one unit. the p-values indictes that relationship between GDP and life expectancy varies depending on the level of total expenditure.


#Hypothesis Testing using ANOVA 

```{r }
#Null Hypothesis: GDP and Total Expenditure do not significantly affect life expectancy.
#Alternative Hypothesis: At least one of GDP or Total Expenditure significantly affects life expectancy. 
anova_result <- aov(life_expectancy ~ total_expenditure * gdp, data = WHO)
# View the summary of the ANOVA results
summary(anova_result)
```

#Explanation of hypothesis testing using ANOVA

#Evaluating each factor
total_expenditure: The F value of 63.45 indicates that total expenditure has a significant effect on life expectancy.
The p-value (3.05e-15) is much less than 0.05, indicating that you can reject the null hypothesis for total expenditure. This means total expenditure significantly affects life expectancy.
gdp: The F value of 360.79 suggests a very strong effect of GDP on life expectancy. The p-value (< 2e-16) is extremely small, leading to the rejection of the null hypothesis for GDP as well. This indicates that GDP significantly affects life expectancy.
total expenditure:gdp: The F value of 9.18 indicates a significant interaction between total expenditure and GDP regarding their combined effect on life expectancy. The p-value (0.00248) is less than 0.01, suggesting that the interaction effect is statistically significant. This means that the effect of total expenditure on life expectancy is influenced by the level of GDP.

Meaning: Both total expenditure and GDP have significant effects on life expectancy. There is a significant interaction effect between total expenditure and GDP, meaning that the impact of total expenditure on life expectancy varies depending on the level of GDP. These results all suggest that we can reject the null hypothesis that GDP and Total Expenditure do not significantly affect life expectancy.


# Visualizing the Interaction
```{r}

ggplot(WHO, aes(x = gdp, y = life_expectancy, color = cut(total_expenditure, 5))) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE)
```

The figure shows that the the effect of GDP on the life expectancy really depends on the total expenditure value 



# Visualizing the Interaction based on the value of total_expenditure
```{r}

# Create a new variable 'expenditure_group' by cutting 'total_expenditure' into 5 groups
WHO$expenditure_group <- cut(WHO$total_expenditure, 5)

ggplot(WHO, aes(x = gdp, y = life_expectancy, color = expenditure_group)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE)+
  facet_wrap(~ expenditure_group) # create separate panels for each level of 'expenditure_group' 
```


# Visualizing the Interaction based on the stauts of the country and expenditure_group
```{r}


ggplot(WHO, aes(x = gdp, y = life_expectancy, color = expenditure_group)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(~ status + expenditure_group) # Create separate panels for each level of 'expenditure_group' and the country status
```

# confedence interval for for coefficient the multi_model_gdp_status Model

```{r}
 
 confint(multi_model_gdp_status)

```

1- We can see that the confidence interval for the intercept ranges from 6.420144e+01 to 6.630828e+01. This means that we can be 95% confident that the true value of the intercept lies within this range.

2- Similarly, the confidence interval for GDP is between 2.910889e-04 and 3.583137e-04, indicating that we can be 95% confident that the true effect of GDP on life expectancy falls within this interval.

3- According to total expenditure, the confidence interval for total expenditure is between 2.083665e-01 and 5.438757e-01, indicating that we can be 95% confident that the true effect of total expenditure on life expectancy falls within this interval.

Since the confidence intervals for the intercept, GDP, and total expenditure do not include zero, we can reject the null hypothesis, which suggests that there is a statistically significant relationship between GDP, total expenditure, and life expectancy.




# visualizing confedence interval for for coefficient the multi_model_gdp_status Model


```{r }

# function to visualize confidence interval for for coefficient the multi_model_gdp_status Model
visualize_coefficient_CI <- function(function_model) {
  # Create data frame for coefficients
  df_coefficient <- as.data.frame(confint(function_model))
  df_coefficient$coefficient <- rownames(df_coefficient)
  df_coefficient$estimate <- coef(function_model)

  ggplot(df_coefficient, aes(x = coefficient, y = estimate)) +  
    geom_point(size = 2) +
   geom_errorbar(aes(ymin = `2.5 %`, ymax = `97.5 %`), width = 0.2) +

    labs(title = "Coefficients and Their Confidence Intervals", 
         x = "Coefficient", 
         y = "Estimate")
}
visualize_coefficient_CI(multi_model_gdp_status)

```

1- Confidence Interval for **intercept**seems to be slightly small. This indicates that the intercept is not very precise estimated..

Similarity, the Confidence Interval for GDP and total expentiture are tiny. This indicates that the 
GDP is precise estimated.



# visualizing multi_model_gdp_status Model
```{r}

#Function to visualize multi_model_gdp_status Model
visualize_multiple_regression <- function(dependent_name, independent_vars, independent_names) {
  
  # Create a ggplot
  ggplot(WHO, aes_string(x = independent_vars, y = dependent_name)) + 
    geom_point() + # visualize the data points
    geom_smooth(method = "lm", se = FALSE, color = "blue") + # # Add a linear regression line 
    theme_minimal() +
    labs(title = paste("Visualizing : ", dependent_name, " vs ", independent_names),
         x = independent_names,
         y = dependent_name)
}

# Calling visualize_multiple_regression function 
visualize_multiple_regression("life_expectancy",  # the dependent variable name
                               "gdp + total_expenditure",# passing the independent variables as a string
                               "GDP && Total Expenditure") # column names to be shown in the figure
     

```
# visualzing the residuals 

```{r }
library(ggplot2)

# Function to visualize residuals
visualize_residuals <- function(dependent_name, independent_vars, independent_names, data_model) {
  
  # Predict values from the model and add them to the WHO dataset
  WHO$predicted <- predict(data_model)
 
  # Plotting
  ggplot(WHO, aes_string(x = independent_vars, y = dependent_name)) +
    geom_point() + # actual data
    
    ## add red segments to represent the residual
    geom_segment(aes_string(xend = independent_vars, yend = "predicted"), color = "red") + 
    
    # add a linear regression line
    geom_smooth(method = "lm", se = FALSE, color = "blue")+  
    theme_minimal() + 
    labs(title = paste("Visualizing Residuals: ", dependent_name, " vs ", independent_names),
         x = independent_names,
         y = "Residuals")
}


#WHO$log_life_expectancy <- log(WHO$life_expectancy)

visualize_residuals("life_expectancy", # dependent variable name
                    "total_expenditure + gdp",  #Independent variables as a string
                    "GDP per Capita && total Expenditure " # Display name for independent variables
                    ,multi_model_gdp_status ) # The model to use for prediction



```

1- The plot shows some points with very large residuals, especially on the right side and the left side these are likely outlier. 



```{r }


# Function to visualize the confidence interval of a multiple regression model
visualize_confidence_interval_multiple <- function(function_model, dependent_name, independent_vars, independent_names) {
  
   # Create data frame with 200 rows as independent variables
  new_data <- data.frame(matrix(ncol = length(independent_vars), nrow = 200))
  
  # sssign column names to the new data based on independent variable names
  colnames(new_data) <- independent_vars
  
  ##lLoop through each independent variable to 
  for (i in 1:length(independent_vars)) {
    
    
     # Fill each column with spaced values between the min and max of the  each variable
    new_data[[independent_vars[i]]] <- seq(min(WHO[[independent_vars[i]]]), 
                                             max(WHO[[independent_vars[i]]]), 
                                             length.out = 200)

  }
  
  # Predict with confidence interval
  CI_95 <- predict(function_model, newdata = new_data, interval = "confidence", level = 0.95)
  
  # Add fitted , upper, and lower confidence intervals
  new_data$fit <- CI_95[, "fit"]
  new_data$up <- CI_95[, "upr"]
  new_data$down <- CI_95[, "lwr"]
  
  
  # combining the independent variable names
   colnames_format= paste(independent_vars[1] ," + ",independent_vars[2])

  # Create the ggplot
  ggplot(new_data, aes_string(x = colnames_format  , y = "fit")) +  
    geom_line(color = "red") +  # # red line for the fitted values
    geom_ribbon(aes(ymin = down, ymax = up), alpha = 0.3, fill = "green") +  # green ribbon for the confidence interval
    theme_minimal() +
    labs(title = paste("95% confidence Interval for ", independent_names, " vs ", dependent_name),
         x = independent_names,
         y = dependent_name) 
}

visualize_confidence_interval_multiple(multi_model_gdp_status, # passing model name
                                       "life_expectancy", # passing dependent variable name 
                                        c("gdp", "total_expenditure"), # passing independent variables as a vector 
                                        "GDP & total Expenditure ") # # column names to be shown in the figure

```

#Model Diagnostics
```{r }
#Q-Q plot
qqnorm(resid(multi_model_gdp_status), main = "Q-Q Plot for Normality of Residuals")
qqline(resid(multi_model_gdp_status), col = "red")
#Residuals vs. Fitted
plot(multi_model_gdp_status, which=1)
#Scale-Location
plot(multi_model_gdp_status, which=3)
#Cook's Distance
plot(multi_model_gdp_status, which=4)

```

#Analysis of Model Diagnostics

Q-Q plot: The Q-Q plot shows that the residuals follow the red line fairly closely, representing a normal distribution, which is an assumption of linear regression.
Residuals vs. Fitted: In the residuals vs. fitted plot, the red line is not flat, the red line is curved, which suggests that the relationship between the independent and dependent variables might not be well-captured by the model, meaning there could be non-linearity. 
Scale-Location: The scale-location plot shows the red line is upward slanting and the residuals are clustered at one end of the plot and become more spread out as the plot moves along. The scale-location plot suggests heteroscedasticity because the red line is not flat but shows a distinct trend, indicating that the variance of the residuals is not constant across fitted values.
Cook's Distance: None of the observations reach a Cook's distance of 1, so there are no significant observations that are influencing the model. However, it is important to note the spike of observations at around 800, because this data could be affecting the model because of the volume of observation numbers.#Model Diagnostics.



```{r }
#Q-Q plot
qqnorm(resid(multi_model_gdp_status), main = "Q-Q Plot for Normality of Residuals")
qqline(resid(multi_model_gdp_status), col = "red")
#Residuals vs. Fitted
plot(multi_model_gdp_status, which=1)
#Scale-Location
plot(multi_model_gdp_status, which=3)
#Cook's Distance
plot(multi_model_gdp_status, which=4)

```

#Analysis of Model Diagnostics

Q-Q plot: The Q-Q plot shows that the residuals follow the red line fairly closely, representing a normal distribution, which is an assumption of linear regression.
Residuals vs. Fitted: In the residuals vs. fitted plot, the red line is not flat, the red line is curved, which suggests that the relationship between the independent and dependent variables might not be well-captured by the model, meaning there could be non-linearity. 
Scale-Location: The scale-location plot shows the red line is upward slanting and the residuals are clustered at one end of the plot and become more spread out as the plot moves along. The scale-location plot suggests heteroscedasticity because the red line is not flat but shows a distinct trend, indicating that the variance of the residuals is not constant across fitted values.
Cook's Distance: None of the observations reach a Cook's distance of 1, so there are no significant observations that are influencing the model. However, it is important to note the spike of observations at around 800, because this data could be affecting the model because of the volume of observation numbers.


#Effect size

``` {r}
library(effectsize)
library(afex)
library(lme4)
library(Matrix)# lme4 package and Matrix


# https://stats.stackexchange.com/questions/266003/interpretation-of-cohens-f2-for-effect-size-in-multiple-regression


# Fit the multiple linear regression model with both predictors
multi_model_gdp_status <- lm(life_expectancy ~ gdp + total_expenditure, data = WHO)


# Get R-squared values for the model
r_squared_all <- summary(multi_model_gdp_status)$r.squared

# Fit models for individual predictors
model_test_total_expenditure <- lm(life_expectancy ~ total_expenditure, data = WHO)

# Get R-squared values for the model
r_squared_mod_total_expenditure <- summary(model_test_total_expenditure)$r.squared

model_test_gdp <- lm(life_expectancy ~ gdp, data = WHO)
# Get R-squared values for the model
r_squared_mod_gdp <- summary(model_test_gdp)$r.squared

# Calculate Cohen's f^2 for the predictors
cohens_f_gdp <- (r_squared_all - r_squared_mod_gdp) / (1 - r_squared_all)
cohens_f_total_expenditure <- (r_squared_all - r_squared_mod_total_expenditure) / (1 - r_squared_all)

# Print results
#Cohen's f^2  for GDP:
cohens_f_gdp
#Cohen's f^2 for Total Expenditure

cohens_f_total_expenditure


cohens_f(multi_model_gdp_status) # Using afex package to show Cohen's f


```

For total_expenditure, Cohen's f^2 = 0.0117493, which indicates a small effect.
For cohens_f^2 for gdp it equals 0.2181094, which falls in the medium to large effect range.

